{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-05-29 ms.tex
Basic parameters vignette
Comments and edits to the first two sections

\newcommand{\plant}{\textsc{plant}}

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait-based demography, 
  ecology and evolution in vegetations.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into metapopulation level estimates of
  reproductive fitness and vegetation structure.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
% todo
  We provide worked examples illustrating {\plant}'s use to study the
  influence of functional traits on growth of individual plants, of
  whole patches and on assembly of ecological communities.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and trait-
structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. As an individual increases its leaf
area, it increases its potential to generate photosynthate.  In
contrast, as individuals grow, they must use increasing fractions of
this photosynthetic income to building and maintaining support tissues
rather than to new leaves \citep{Givnish-1988, Enquist-2007} and
reproduction \citep{Thomas-2011}. Consequently rates of growth and
mortality change with individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
The patterns of changing growth rate vary by plant traits; leaf
construction costs and wood density affect growth rates and rates of
tissue replacement \citep{Wright-2010}, while seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}
Strong feedbacks emerge between individuals within a forest due to
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals.  This makes it difficult to predict demographic processes
without modelling both individual growth and the interactions between
individuals. The non-linear feedbacks arising from competitive 
interactions are thought to be crucial for understanding phenomena such 
as self thinning, successional transitions, trait evolution and 
species coexistence.

Although the importance of size-structure for vegetation dynamics and diversity, 
has been long been recognised
\citep[e.g.,][]{Harper-1977, Shugart-1980, Huston-1987}, current research in these
areas is dominated by models and theory that either omit size-structure 
entirely, or only include coarse approximations. Theoretical investigations 
exploring questions about niche-based species coexistence and ecological drift 
(neutrality) both primarily rely  on unstructured population models 
\citep[e.g.][]{MacArthur-1967, Tilman-1985, Geritz-1995, Hubbell-2001,
Calcagno-2006, Leimar-2013}. It is common to assume
competitive interactions are influenced by size-related traits such as
adult size and seed size, but detailed size-structured demography is
rarely considered, at least for plant models \cite[for animal
examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has been
difficult to connect and test the mechanisms described in these simplistic models
with types of data collected in the field, such as physiological
leaf traits or demographic data from long-term monitoring plots \citep{Adler-2013}.
Similarly, most large-scale models of vegetation dynamics do not include
size-structured demography \citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. 
During the  1980s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation of models 
focussed more heavily on biogeochemistry at the expense of demography 
\citep[e.g.][]{Woodward-1995,Haxeltine-1996}. While the importance
of size-structure is starting again to be recognised \citep{Moorcroft-2001, 
 Falster-2011, Dekauwe-2014}, all but a couple \citep[e.g.][]{Moorcroft-2001, Smith-2014} 
of the most popular, large-scale vegetation models lack any real demography or 
size-structure.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime. In the theoretical ecology literature it is common to
analyse the full behaviour of a model under its equilibrium conditions.
{\plant} was developed with this style of analysis in mind, while also 
allowing for a richer set of ecological dynamics than is possible in 
population models lacking size structure. {\plant} is thus part
of an extended thought experiment investigating how species differing
in traits are able to coexist with one another \citep{Falster-2011,Falster-2015}.
Understanding and incorporating the demographic outcomes of trait variation 
is ultimately needed if we are to properly understand how trade-offs in plant function 
enable species coexistence.

At the same time, we expect the {\plant} model to be useful for modelling
other  phenomena, beyond questions about species coexistence. The internal
structure  of the {\plant} model is superficially similar to the
\textsc{Ecosystem Demography}  \citep{Moorcroft-2001}  and \textsc{LPJ-GUESS}
\citep{Smith-2014} models. Like  {\plant}, these models include size-
structured competition within a metapopulation of  "patches". However,
whereas these models were designed to capture large-scale regional
biogeochemistry driven by climatic variables and rely on simulations,
{\plant} was designed so that the rules  driving the dynamics of the system
could be manipulated and emergent demographic  behaviours studied in detail.
In {\plant} it is possible to grow plants in isolation, in patches of
competing plants and in disturbance-driven metapopulations. The package is
thus designed to sit in the middle ground between a large-scale coupled
simulation model and simple abstract theoretical device. As examples, we
expect the package will prove  useful for investigating how morphological and
physiological traits influence growth rate and shade tolerance, and emergent
phenomena such as multi-species self-thinning,  successional transitions and
age-related changes in vegetation productivity.

Below we describe the general approach of the package, then provide  sections
with details at different levels of ecological organisation.  Detailed
technical documents are provided as supplementary information  (see
Appendices), and updated versions of technical documents are also available
within the {\plant} package itself.

@@

{\plant} is a mechanistic model, meaning the dynamics of the system
arise from rules specifying how individuals grow and interact.  The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment. Being driven by traits, the model can be extended to potentially 
very many species. Users can also  add their own physiological models.

@@

The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011}, but with extensions allowing for
diameter growth also to be estimated. To properly model trait
variation, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed tradeoffs. For example, the trait \textsc{lma} is used to estimate the
rate of leaf turnover, based on observed scaling relationship spanning
across diverse vegetation and plant functional global
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

@@

Our approach for modelling size-structured population dynamics
is based on the Escalator Boxcar Train technique (\textsc{EBT})
\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{EBT} solves the \textsc{pde}
describing development of \(n\) by approximating the density function
with a collection of cohorts spanning the size spectrum. Following a
disturbance, a series of cohorts are introduced into each patch. These
cohorts are then transported along the characteristic equations of the
\textsc{pde}. Biologically these are the growth trajectories of individuals,
conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
growth and mortality combine to alter the density of individuals along
the trajectories (Fig. \ref{fig:patch}b).

We introduce two extensions to the \textsc{EBT}: i) a new method for estimating
the size-density distribution and ii) a technique for handling strong
size-asymmetric competitive feedbacks, such as occur under strong
competition for light (see Appendix \ref{sec:demography} for
details). The original \textsc{EBT} \citep{Deroos-1988,
  Deroos-1992, Deroos-1997} proceeds by approximating the first and
second moments of the density distribution \(n\) within each
cohort. In {\plant} we use a different approach. Instead of tracking
first and second moments of the size-distribution within cohorts, we
track a point mass estimate of \(n\) along the trajectory for the
boundary of each cohort. We found this approach preferable because it
does not require us to maintain a separate set of equations for the
smallest cohort \citep{Deroos-1997}.

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original \textsc{EBT}, cohorts were introduced at a
fixed time interval. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
\ref{sec:patch} and \ref{sec:cohort-spacing}). This can become a
problem later during patch development because much of the population
can be located between widely-spaced cohorts, delivering low numerical
precision. To maintain numerical precision we therefore use an
iterative algorithm to adaptively refine the cohort spacing until the
trajectories of adjacent cohorts remain adequately resolved throughout
development of the patch (see Appendix \ref{sec:demography} for
details). This refinement causes an uneven distribution of cohort
introduction times across patch age (Fig.  \ref{fig:patch}a), with
considerably tighter spacing of cohorts in younger patches.

@@

The main numerical challenge in stepping from single patch to
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An "input" seed rain is required to simulate 
the demography of the metapopulation, this in turn produces an "output" 
number of seeds. The equilibrium seed rain is the vector of seed 
rain where input seed
rain equals output seed rain. Finding this equilibrium requires
finding a root of a multidimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable) (Appendix
\ref{sec:equilibrium}).

@@

Beyond its conceptual appeal, this idea of a dynamic equilibrium
allows us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits,
competitive interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of
patches. Unlike some vegetation models, there is no need to set the
density of seeds or seedlings, these arise as outputs from the
model. Characteristics of the vegetation such as total leaf area,
biomass, productivity and allocation to different plant parts also
arise as emergent properties (Fig.  \ref{fig:patch}c). Similarly,
distributions of abundance, leaf area and growth rate with respect to
size emerge at both patch scale and scale of metapopulation
(Fig. \ref{fig:emergent}). Note in particular how a single
physiological model produces a variety of height-density
relationships, dependent on patch age, as well as an emergent average
across all patches (Fig. \ref{fig:emergent}, see more detail in
appendix \ref{sec:emergent}). This emergent relationship might then be
compared to the observed distributions in large forest plots
\citep{Muller-2006}.

@@

Fitness landscapes will be positive in regions that allow invasion,
and zero for species that are at equilibrium. The slope of the fitness
landscape at a resident indicates the direction of selection; a
negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
selection for smaller trait values (see appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution
\citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient,
can lead to an evolutionary branching point
\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
directional selection (solid line in Fig. \ref{fig:fitness}a) but
invasion is possible both above and below it in trait space. It is
often possible to find a configuration where fitness is negative
everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
points where species are present. This community is immune to invasion
from any other strategies \citep{Geritz-1998}.

Identifying conditions promoting trait diversity is a valuable use-case
for {\plant}. Coexisting plant species differ in a range of physiological
traits, yet the conditions allowing for these different types to coexist
in face of competition for a common set of resources remain largely
unknown. The simple example here can be extended to multiple dimensions
and can be used to investigate how tradeoffs facilitate coexistence
\citep{Falster-2015}

@@

{\plant} makes use of much existing software, including the programming
language C++; the R computational environment \citep{R-2015}, the R
packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
\texttt{R6} \citep{Chang-2014} and \texttt{testhat}
\citep{Wickham-2011}; and the Boost Library for C++
\citep{Schaling-2014} via the R package \texttt{BH}
\citep{Eddelbuettel-2015}. Source code is hosted at
\href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
Installation instructions are available at this homepage, binary
releases will also be available there.  (Note that {\plant} does not
currently compile on Windows, but this will change soon as R core
upgrade the Windows C++ compiler).

@@

\caption{Processes modelled within {\plant} include physiological, population and
evolutionary dynamics.
Dynamics across all three levels are driven by the
physiological sub model determining an individual's physiological and
demographic rates on the basis of its traits, size and light environment (a).
(b) Competitive hierarchies  are modelled by tracking the height distribution of
individuals across within a "patch" after a disturbance. The intensity
of shading indicates the density of individuals at a given height for
different species (distinguished by colours). A size- and patch-structured
metapopulation consists of a distribution of patches linked by seed dispersal.
Disturbances occasionally remove all vegetation within a patch, resetting the
patch. (c) The traits of the resident species determine the shading environment
across the metapopulation, which in turn determines the fitness landscape.
Resident traits adjust through directional selection up the fitness landscape
and through the introduction of new species. Adapted from
\citet{Falster-2011}, \citet{Falster-2015}. }

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
\caption{Population dynamics for two species competing within a patch. Lines
in panels (a) same growth trajectories for individuals on the
boundary of successive "cohorts" for each species, using a schedule of 
introduction times generated by an adaptive refinement algorithm.
 Red lines are the low \textsc{lma}
species from figure \ref{fig:plant}, blue lines are the high \textsc{lma} species.  The
initial growth rate advantage of the low \textsc{lma} species (see Figure
\ref{fig:plant} c) means it quickly over tops the high \textsc{lma} species,
suppressing its growth.  Self-thinning leaves space for the high \textsc{lma} species
to establish below the canopy of the fast species. Panel (b) shows the same
trajectories as in panel (a), but with shading indicating the density $n$ of
individuals at given size and patch age (light = low density, dark = high density).  
(c) The leaf area index at ground
level for each species.  The dotted line is the total leaf area index (sum
across both species). See Appendix \ref{sec:patch} for details and code.}
\label{fig:patch}
\end{figure}

@@

\section{Modifying parameters of physiological model}\label{sec:parameters}

See attached file \url{vignette_parameters.R}

@@

>>>>

\newcommand{\plant}{\texttt{plant}}

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait-based demography, 
  ecology and evolution in simulated forests.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into metapopulation level estimates of
  reproductive fitness and vegetation structure.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
% todo
  We provide worked examples illustrating {\plant}'s use to study the
  influence of functional traits on growth of individual plants, of
  whole patches and on assembly of ecological communities.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and trait-
structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. 
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
In opposition to this, as individuals grow they must use increasing
fractions of their photosynthetic income to building tissues other
than new leaves; to maintaining support tissues \citep{Givnish-1988,
  Enquist-2007} and to reproduction \citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  The
patterns of changing growth rate vary with plant traits; for example,
leaf construction costs and wood density affect growth rates and rates
of tissue replacement \citep{Wright-2010}, while seed size and height
at maturation determine start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest due to
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals.
%
This makes it difficult to predict demographic processes without
modelling both individual growth and the interactions between
individuals.
%
It is therefore difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence.

Although the importance of size-structure for vegetation dynamics and
diversity, has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations. 
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both primarily
rely on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive interactions are influenced by
size-related traits such as adult size and seed size, but detailed
size-structure is rarely considered, at least for plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
% TODO: I don't have the background to write it, but (at the risk of
% making a long paragraph even longer) is it worth adding something
% saying something about how we'd expect size structure to be
% particularly important for plants because they grow up through a
% canopy?  Perhaps not useful?
Consequently, it has been difficult to connect and test the mechanisms
described in these simplistic models with types of data collected in
the field, such as physiological leaf traits or demographic data from
long-term monitoring plots \citep{Adler-2013}.
%
Similarly, most large-scale models of vegetation dynamics do not
include size-structured demography \citep{Cramer-2001, Sitch-2003,
  Kelley-2013,Dekauwe-2014}.  During the 1980s, size-structured
``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation
of models focused more heavily on biogeochemistry at the expense of
demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996}. While the
importance of size-structure is starting again to be recognised
\citep{Moorcroft-2001, Falster-2011, Dekauwe-2014}, all but a couple
\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular,
large-scale vegetation models lack any real demography or
size-structure.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
% TODO: the first sentence reads weirdly to me: the paragraph now says
% "we introduce our model / other models are like X / our model is
% also like X" I'm not sure what the key point that is being driven at
% here, either.
%
% TODO: I've reordered it and changed the emphasis here.  Can you
% check that you agree and if so add a reference.
We developed {\plant} primarily to investigate if and how species
differing in traits are able to coexist with one another
\citep{Falster-2011,Falster-2015}.  In order to allow interactions to
be driven by species traits and arise through competition for shared
resources, we needed to model a community with a full range of sizes,
as species may not be able to recruit under the canopy that they
generate (REF).

At the same time, we expect the {\plant} model to be useful for
modelling other phenomena. The internal structure of the {\plant}
model is similar in approach to the ``Ecosystem Demography''
\citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
models. Like {\plant}, these models include size- structured
competition within a metapopulation of ``patches''. However, whereas
these models were designed to capture large-scale regional
biogeochemistry driven by climatic variables and rely on simulations,
{\plant} was designed so that the rules driving the dynamics of the
system could be manipulated and emergent demographic behaviours
studied in detail.  In {\plant} it is possible to grow plants in
isolation, in patches of competing plants and in disturbance-driven
metapopulations. The package is thus designed to sit in the middle
ground between a large-scale coupled simulation model and simple
abstract theoretical device. 
% TODO: Consider moving to the conclusion?
As examples, we expect the package will prove useful for investigating
how morphological and physiological traits influence growth rate and
shade tolerance, and emergent phenomena such as multi-species
self-thinning, successional transitions and age-related changes in
vegetation productivity.

% TODO: they aren't actually available within plant yet, and
% realistically won't be by the updated.  I've added a placeholder
% that contains appropriate links.
Below we describe the general approach of the package, then provide
sections with details at nested levels of ecological organisation.
Detailed technical documents are provided as supplementary information
(see Appendices), and updated versions of technical documents are also
available within the {\plant} package itself.

@@

{\plant} is a mechanistic model, meaning the dynamics of the system
arise from rules specifying how individuals grow and interact.  The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment. Being driven by traits, the model
can be extended to potentially very many species. Users can also add
their own physiological models, though this is a work in progress.

@@

The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011}, but with extensions allowing for
diameter growth also to be estimated. To properly model trait
variation, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed trade-offs. For example, the trait \textsc{lma} is used to estimate the
rate of leaf turnover, based on observed scaling relationship spanning
across diverse vegetation and plant functional global
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

@@

Our approach for modelling size-structured population dynamics
is based on the Escalator Boxcar Train technique (\textsc{ebt})
\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves the \textsc{pde}
describing development of \(n\) by approximating the density function
with a collection of cohorts spanning the size spectrum. Following a
disturbance, a series of cohorts are introduced into each patch. These
cohorts are then transported along the characteristic equations of the
\textsc{pde}. Biologically these are the growth trajectories of individuals,
conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
growth and mortality combine to alter the density of individuals along
the trajectories (Fig. \ref{fig:patch}b).

We introduce two extensions to the \textsc{ebt}: i) a new method for estimating
the size-density distribution and ii) a technique for handling strong
size-asymmetric competitive feedbacks, such as occur under strong
competition for light (see Appendix \ref{sec:demography} for
details). The original \textsc{ebt} \citep{Deroos-1988,
  Deroos-1992, Deroos-1997} proceeds by approximating the first and
second moments of the density distribution \(n\) within each
cohort. In {\plant} we use a different approach. Instead of tracking
first and second moments of the size-distribution within cohorts, we
track a point mass estimate of \(n\) along the trajectory for the
boundary of each cohort. We found this approach preferable because it
does not require us to maintain a separate set of equations for the
smallest cohort \citep{Deroos-1997}.

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original \textsc{ebt}, cohorts were introduced at a
fixed time interval. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
\ref{sec:patch} and \ref{sec:cohort-spacing}). This can become a
problem later during patch development because much of the population
can be located between widely-spaced cohorts, delivering low numerical
precision. To maintain numerical precision we therefore use an
iterative algorithm to adaptively refine the cohort spacing until the
trajectories of adjacent cohorts remain adequately resolved throughout
development of the patch (see Appendix \ref{sec:demography} for
details). This refinement causes an uneven distribution of cohort
introduction times across patch age (Fig.  \ref{fig:patch}a), with
considerably tighter spacing of cohorts in younger patches.

@@

The main numerical challenge in stepping from single patch to
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An ``input'' seed rain is required to simulate 
the demography of the metapopulation, this in turn produces an ``output'' 
number of seeds. The equilibrium seed rain is the vector of seed 
rain where input seed
rain equals output seed rain. Finding this equilibrium requires
finding a root of a multidimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable) (Appendix
\ref{sec:equilibrium}).

@@

Beyond its conceptual appeal, this idea of a dynamic equilibrium
allows us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits,
competitive interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of
patches. Unlike some vegetation models, there is no need to set the
density of seeds or seedlings, these arise as outputs from the
model. Characteristics of the vegetation such as total leaf area,
biomass, productivity and allocation to different plant parts also
arise as emergent properties (Fig.  \ref{fig:patch}c). Similarly,
distributions of abundance, leaf area and growth rate with respect to
size emerge at both patch scale and scale of metapopulation
(Fig. \ref{fig:emergent}). Note in particular how a single
physiological model produces a variety of height-density
relationships, dependent on patch age, as well as an emergent average
across all patches (Fig. \ref{fig:emergent}, see more detail in
Appendix \ref{sec:emergent}). This emergent relationship might then be
compared to the observed distributions in large forest plots
\citep{Muller-2006}.

@@

Fitness landscapes will be positive in regions that allow invasion,
and zero for species that are at equilibrium. The slope of the fitness
landscape at a resident indicates the direction of selection; a
negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
selection for smaller trait values (see Appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution
\citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient,
can lead to an evolutionary branching point
\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
directional selection (solid line in Fig. \ref{fig:fitness}a) but
invasion is possible both above and below it in trait space. It is
often possible to find a configuration where fitness is negative
everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
points where species are present. This community is immune to invasion
from any other strategies \citep{Geritz-1998}.

Identifying conditions promoting trait diversity is a valuable use-case
for {\plant}. Coexisting plant species differ in a range of physiological
traits, yet the conditions allowing for these different types to coexist
in face of competition for a common set of resources remain largely
unknown. The simple example here can be extended to multiple dimensions
and can be used to investigate how trade-offs facilitate coexistence
\citep{Falster-2015}

@@

{\plant} makes use of much existing software, including the programming
language C++; the R computational environment \citep{R-2015}, the R
packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
\texttt{R6} \citep{Chang-2014} and \texttt{testthat}
\citep{Wickham-2011}; and the Boost Library for C++
\citep{Schaling-2014} via the R package \texttt{BH}
\citep{Eddelbuettel-2015}. Source code is hosted at
\href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
Installation instructions are available at this homepage, binary
releases will also be available there.  (Note that {\plant} does not
currently compile on Windows, but this will change soon as R core
upgrade the Windows C++ compiler).

@@

\caption{Processes modelled within {\plant} include physiological, population and
evolutionary dynamics.
Dynamics across all three levels are driven by the
physiological sub model determining an individual's physiological and
demographic rates on the basis of its traits, size and light environment (a).
(b) Competitive hierarchies  are modelled by tracking the height distribution of
individuals across within a ``patch'' after a disturbance. The intensity
of shading indicates the density of individuals at a given height for
different species (distinguished by colours). A size- and patch-structured
metapopulation consists of a distribution of patches linked by seed dispersal.
Disturbances occasionally remove all vegetation within a patch, resetting the
patch. (c) The traits of the resident species determine the shading environment
across the metapopulation, which in turn determines the fitness landscape.
Resident traits adjust through directional selection up the fitness landscape
and through the introduction of new species. Adapted from
\citet{Falster-2011}, \citet{Falster-2015}. }

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
\caption{Population dynamics for two species competing within a patch. Lines
in panels (a) same growth trajectories for individuals on the
boundary of successive ``cohorts'' for each species, using a schedule of 
introduction times generated by an adaptive refinement algorithm.
 Red lines are the low \textsc{lma}
species from Fig. \ref{fig:plant}, blue lines are the high \textsc{lma} species.  The
initial growth rate advantage of the low \textsc{lma} species (see Figure
\ref{fig:plant} c) means it quickly over tops the high \textsc{lma} species,
suppressing its growth.  Self-thinning leaves space for the high \textsc{lma} species
to establish below the canopy of the fast species. Panel (b) shows the same
trajectories as in panel (a), but with shading indicating the density $n$ of
individuals at given size and patch age (light = low density, dark = high density).  
(c) The leaf area index at ground
level for each species.  The dotted line is the total leaf area index (sum
across both species). See Appendix \ref{sec:patch} for details and code.}
\label{fig:patch}
\end{figure}

@@

\section{Modifying parameters of the physiological model}\label{sec:parameters}

See attached file \url{vignettes/parameters.pdf}, knitr source
\url{vignettes/parameters.Rmd}, plain source, \url{vignettes/R/parameters.R}

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% TeX-PDF-mode: t
%%% End:

%  LocalWords:  Westoby photosynthate Givnish Enquist Ruger feedbacks
%  LocalWords:  ontogenetic Shugart Tilman Geritz Hubbell Calcagno BH
%  LocalWords:  Cramer Sitch Dekauwe focussed successional Haxeltine
%  LocalWords:  Moorcroft lpj FFW lma wplcp Baltzer Lusk Kohyama pde
%  LocalWords:  Deroos Pacala Strigul ebt adaptively Coomes Binkley
%  LocalWords:  Ogawa Bormann Gyllenberg Vonfoerster Metz Dieckmann
%  LocalWords:  Rcpp Eddelbuettel templated RcppR testhat Wickham DS
%  LocalWords:  Schaling RG Brn nnstr Camac De Roos Duursma Johansson
%  LocalWords:  knitr
